---
title: "R Notebook"
output: html_notebook
---

Load the required packages
```{r}
suppressMessages(library("MultiAssayExperiment"))
suppressMessages(library("MOFA"))
suppressMessages(library("MOFAdata"))
#install.packages("r.jive")
suppressMessages(library("r.jive"))
#BiocManager::install("omicade4")
suppressMessages(library("omicade4"))
#BiocManager::install("BloodCancerMultiOmics2017")
suppressMessages(library("BloodCancerMultiOmics2017"))
suppressMessages(library("survival"))
suppressMessages(library("survminer"))
suppressMessages(library("zoo"))
suppressMessages(library("pheatmap"))
suppressMessages(library("fgsea", quietly = TRUE))
#BiocManager::install("biomaRt")
suppressMessages(library("biomaRt"))
```

```{r}
source("./scripts/HelperForTheCourse.R")
```

Load the data
```{r}
data("CLL_data", package = "MOFAdata")
data("CLL_covariates", package = "MOFAdata")
```
CLL_covariates only has gender and disgnosis information

## MOFA

Create MOFA object
```{r}
MOFAobject <- createMOFAobject(CLL_data)
```

```{r}
MOFAobject
```

```{r}
plotDataOverview(MOFAobject)
```

Load pre-trained MOFA model (assuming that training has already been covered in the talk)
```{r}
filepath <- system.file("extdata", "CLL_model.hdf5", package = "MOFAdata")
MOFAobject <- loadModel(filepath, MOFAobject)
```

```{r}
MOFAobject
```
We limit the number of factors to 10 in all the subsequent analysis.
```{r}
numFactors = 10
```

Extract weights and factors matrix
```{r}
MOFAweights <- getWeights(MOFAobject)
```

```{r}
MOFAfactors <- getFactors(MOFAobject)
```

## Impute the data
############# Replacing NAs with feature-wise mode in case of mutation data and feature-wise mean in case of other non-binary data ######################
############# Or remove mutation data as imputing is too hard/dangerous ######################
```{r}
#To make the comparisons with MOFA, fill in the missing values with feature-wise mean
#Participants can play around here with various imputation techniques
CLL_data_imputed <- list("Mutations" = t(apply(CLL_data[["Mutations"]], 1, function(x) {
  x[which(is.na(x))] <- calculateMode(x, na.rm = TRUE)
  return(x)})))
CLL_data_imputed[c("Drugs", "mRNA", "Methylation")] <- lapply(CLL_data[c("Drugs", "mRNA", "Methylation")], function(x) t(na.aggregate(t(x))))
```

# Note that the following sections are derived from the momix pipeline made available by Laura et al. (2020)

## JIVE

```{r}
JIVEfactorization <- jive(CLL_data_imputed, rankJ=numFactors, rankA = rep(numFactors, length(CLL_data_imputed)), method = "given", conv = "default", maxiter = 100, showProgress=TRUE)
```

```{r}
rankJV <- JIVEfactorization$rankJ;
rankIV.v <- JIVEfactorization$rankA;
J <- numeric(0)
ng <- 0
JIVEweights <- list();
for(j in 1:length(CLL_data_imputed)){
  J <- rbind(J,JIVEfactorization$joint[[j]]);
  ng <- c(ng,dim(JIVEfactorization$joint[[j]])[1])
}
svd.o <- svd(J);
jV <- svd.o$v %*% diag(svd.o$d);
for(j in 1:length(CLL_data_imputed)){
  JIVEweights[[j]] <- svd.o$u[(1+sum(ng[1:j])):sum(ng[1:j+1]),1:rankJV]; ###error in dimension
  rownames(JIVEweights[[j]]) <- rownames(CLL_data_imputed[[j]])
  colnames(JIVEweights[[j]]) <- paste0("LF", 1:numFactors)
}
names(JIVEweights) <- names(CLL_data_imputed)

```  

```{r}
JIVEfactors <- jV[,1:rankJV]
rownames(JIVEfactors) <- colnames(CLL_data_imputed[[1]])
colnames(JIVEfactors) <- paste0("LF", 1:numFactors)
```
  
## MCIA

```{r}
MCIAfactorization <- mcia(CLL_data_imputed, cia.nf = numFactors)
```

```{r}
MCIAfactors <- as.matrix(MCIAfactorization$mcoa$SynVar)
colnames(MCIAfactors) <- paste0("LF", 1:numFactors)
```

```{r}
MCIAweights <- list()
for(j in 1:length(CLL_data_imputed)){
    MCIAweights[[j]]<-as.matrix(MCIAfactorization$mcoa$axis[1:dim(CLL_data_imputed[[j]])[1],])
    rownames(MCIAweights[[j]])<-rownames(CLL_data_imputed[[j]])
    colnames(MCIAweights[[j]])<-paste0("LF", 1:numFactors)
}
names(MCIAweights) <- names(CLL_data_imputed)
```

## Compare the results

```{r}
MOFA::plotVarianceExplained(MOFAobject)
plot(JIVEfactorization)
plot(MCIAfactorization, axes = c(1,2), sample.lab = FALSE, df.color = 1:4)
plot(MCIAfactorization, axes = c(1,3), sample.lab = FALSE, df.color = 1:4)
plot(MCIAfactorization, axes = c(2,3), sample.lab = FALSE, df.color = 1:4)
```
In case of MOFA = Along with variance explained per view, variance explained per factor per view can also be visualized.
In case of JiVE = Only variance explained per view can be visualized
In case of MCIA = factors = PCs, sorted by their eigen-values as we have selected the first 10 eigenvectors for this analysis. The variance explained per factor per view is not available but overall the eigen-vectors are sorted by their eigenvalues.

## Benchmark 1: Clinical association with IGHV and trisomy 12

```{r}
library(pheatmap)
source("./scripts/HelperForTheCourse.R")
visualizeWeightsHeatmap(MOFAweights, 'Mutations')
visualizeWeightsHeatmap(JIVEweights, 'Mutations')
visualizeWeightsHeatmap(MCIAweights, 'Mutations')
```

```{r}
metagenesList = list("MOFA" = MOFAweights,"MCIA" = MCIAweights, "JIVE" = JIVEweights)
```

```{r}
source("./scripts/HelperForTheCourse.R")
plotTopWeightsUsingSeparateWeightsAndFactors(metagenesList[["MCIA"]], "Mutations", factor = 7, nfeatures = 10)
```

```{r}
data(lpdAll, patmeta, drugs, package = "BloodCancerMultiOmics2017")
```

```{r}
survT <- extractMetadata(lpdCLL, patmeta, drugs)
```

```{r}
factorsList = list("MOFA" = MOFAfactors,"MCIA" = MCIAfactors, "JIVE" = JIVEfactors)
factorsList = lapply(factorsList, function(g) {
  indices = match(rownames(survT), rownames(g))
  return(g[indices, ])
})
```

```{r}
ggplot(data.frame(MOFAfactors)[match(rownames(survT), rownames(MOFAfactors)), ], aes(x = LF1, y = LF2, colour = factor(survT$IGHV), shape = factor(survT$trisomy12))) +
  geom_point()
```
```{r}
ggplot(data.frame(JIVEfactors)[match(rownames(survT), rownames(JIVEfactors)), ], aes(x = LF1, y = LF2, colour = factor(survT$IGHV), shape = factor(survT$trisomy12))) +
  geom_point()
```
```{r}
ggplot(data.frame(MCIAfactors)[match(rownames(survT), rownames(MCIAfactors)), ], aes(x = LF1, y = LF3, colour = factor(survT$IGHV), shape = factor(survT$trisomy12))) +
  geom_point()
```

```{r}
ggplot(data.frame(MCIAfactors)[match(rownames(survT), rownames(MCIAfactors)), ], aes(x = LF2, y = LF3, colour = factor(survT$IGHV), shape = factor(survT$trisomy12))) +
  geom_point()
```

## Benchmark 2: Association with biological annotations

```{r}
source("./scripts/HelperForTheCourse.R")
data("reactomeGS")
pval.thr=0.05
#pathToGeneSymbols = "./data/GeneSymbols.tsv"
#databasePath <- "./data/c2.cp.reactome.v7.0.symbols.gmt" #REACTOME
#databasePath <- "./data/h.all.v7.0.symbols.gmt" #Hallmarks
#databasePath <- "./data/c5.all.v7.0.symbols.gmt" #GO
out_bio <- biologicalComparisonModifiedWithMofaCode(metagenesList, factorsList, CLL_data, pathToFseaScript = "./scripts/FSEA.R", database = reactomeGS, pval.thr=pval.thr)
```

```{r}
out_bio
```

```{r}
#tiff("BioEnr_go_fig1.tiff", units="in", width=8.95, height=6.05, res=300)
biological_analysis <- data.frame(
    matrix(data = NA, ncol=4, nrow=0, 
           dimnames = list(c(), c("methods", "selectivity", "nonZeroFacs", "total_pathways"))
          ),
    stringsAsFactors = FALSE)
biological_analysis <- rbind(biological_analysis,
                                data.frame(methods=rownames(out_bio), out_bio))

min_nonZero = min(biological_analysis[, "nonZeroFacs"]) 
max_nonZero = max(biological_analysis[, "nonZeroFacs"]) 
g <- ggplot(biological_analysis, 
            aes(x=nonZeroFacs,y=selectivity)) + 
    geom_point(aes(colour = methods), size=5, alpha=.6, position=position_jitter(h=0, w=0.15))+ 
    theme_bw() + 
    scale_shape_manual(values=c(15,17,16)) + 
    scale_color_manual(values=c('#FF00FF', '#FF6E28', '#C8961E', '#FF0000', '#0000FF', '#A0A0A0', '#48D1CC', '#00FF00')) +
    #ylim(floor(min((biological_analysis[,"selectivity"]*10)-.4)) / 10,
    #     ceiling(max((biological_analysis[,"selectivity"]*10)+.2)) / 10) +
    labs(title="Biological annotations", 
         x="# metagenes (factors) enriched in at least one annotation") +
    theme(plot.title = element_text(size=14,face="bold"),
          axis.text = element_text(size=11),
          axis.title = element_text(size=13),
          legend.text=element_text(size=10)) +
    ylab("Selectivity") + 
    ylim(c(0, 1))+
    labs(colour = "Methods", shape = "Cancer") +
    guides(color = guide_legend(order = 1),size = guide_legend(order = 3)) + 
    #scale_x_discrete() + scale_x_discrete(limits=min_nonZero:max_nonZero) + 
    scale_x_discrete(limits=min_nonZero:max_nonZero, labels = c(min_nonZero:max_nonZero));
g
#dev.off()
```

```{r}
source("./scripts/FSEA.R")
data("reactomeGS")
gseaResults = list()
for (methodName in names(metagenesList)){
  gseaResults[[methodName]] <- runEnrichmentAnalysisUsingSeparateWeightsAndFactors(
    metagenesList[[methodName]], factorsList[[methodName]], CLL_data,
    view = "mRNA",
    feature.sets = reactomeGS,
    alpha = pval.thr)
  
  #plotEnrichmentBars(gsea, alpha=pval.thr)
  
}
```

```{r}
# heatmap of enriched pathways per factor at 1% FDR
# plotEnrichmentHeatmap(gsea, alpha=pval.thr)
# plot number of enriched pathways per factor at 1% FDR
plotEnrichmentBars(gseaResults[["MOFA"]], alpha=pval.thr)
plotEnrichmentBars(gseaResults[["MCIA"]], alpha=pval.thr)
plotEnrichmentBars(gseaResults[["JIVE"]], alpha=pval.thr)
```


```{r}
source("./scripts/FSEA.R")
interestingFactors <- 4:5

fseaplots <- lapply(interestingFactors, function(factor) {
  plotEnrichmentUsingSeparateWeightsAndFactors(
    gseaResults[["MOFA"]],
    factor = factor,
    alpha = pval.thr,
    max.pathways = 10 # The top number of pathways to display
  )
})

cowplot::plot_grid(fseaplots[[1]], fseaplots[[2]],
                   ncol = 1, labels = paste("Factor", interestingFactors))

fseaplots <- lapply(interestingFactors, function(factor) {
  plotEnrichmentUsingSeparateWeightsAndFactors(
    gseaResults[["MCIA"]],
    factor = factor,
    alpha = pval.thr,
    max.pathways = 10 # The top number of pathways to display
  )
})

cowplot::plot_grid(fseaplots[[1]], fseaplots[[2]],
                   ncol = 1, labels = paste("Factor", interestingFactors))

interestingFactors <- c(5, 7)
fseaplots <- lapply(interestingFactors, function(factor) {
  plotEnrichmentUsingSeparateWeightsAndFactors(
    gseaResults[["JIVE"]],
    factor = factor,
    alpha = pval.thr,
    max.pathways = 10 # The top number of pathways to display
  )
})

cowplot::plot_grid(fseaplots[[1]], fseaplots[[2]],
                   ncol = 1, labels = paste("Factor", interestingFactors))
```

## Benchmark 3: Prediction of survival outcomes

```{r}
source("./scripts/HelperForTheCourse.R")
fp(factorsList, survT, "myPlot")
```