---
title: "Session3_2"
author: "Manik Garg"
date: "26/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Benchmark 2: Association with biological annotations

### Load required packages
```{r}
suppressMessages(library("fgsea"))
suppressMessages(library("ggplot2"))
suppressMessages(library("MOFAdata"))
```

### Load the required data
```{r}
# Load weights obtained from each method in Session 2.1 
load("../results/WeightsList.Rdata")
# Load factors obtained from each method in Session 2.1 
load("../results/FactorsList.Rdata")
```

```{r}
# Load the CLL data and the reactome pathway information
data("CLL_data", "reactomeGS", package = "MOFAdata")
```

### Analyze the biological pathway(s) associated with each factor

Perform gene-set enrichment analysis
```{r}
pathToFseaScript = "../scripts/FSEA.R"
source(pathToFseaScript)
pval.thr = 0.01 #FDR cut-off
gseaResults = list()
for (methodName in names(weightsList)){
  gseaResults[[methodName]] <- runEnrichmentAnalysisUsingSeparateWeightsAndFactors(
    weightsList[[methodName]], factorsList[[methodName]], CLL_data,
    view = "mRNA",
    feature.sets = reactomeGS,
    alpha = pval.thr)
}
```

```{r}
# plot number of enriched pathways per factor at 1% FDR
methodName = "MOFA"
plotEnrichmentBars(gseaResults[[methodName]], methodName, alpha=pval.thr)
methodName = "JIVE"
plotEnrichmentBars(gseaResults[[methodName]], methodName, alpha=pval.thr)
methodName = "MCIA"
plotEnrichmentBars(gseaResults[[methodName]], methodName, alpha=pval.thr)
```


```{r}
interestingFactors <- c(5,4)

methodName = "MOFA"
fseaplots <- lapply(interestingFactors, function(factor) {
  plotEnrichmentUsingSeparateWeightsAndFactors(
    gseaResults[[methodName]],
    methodName,
    factor = factor,
    alpha = pval.thr,
    max.pathways = 10 # The top number of pathways to display
  )
})

cowplot::plot_grid(fseaplots[[1]], fseaplots[[2]],
                   ncol = 1, labels = paste("Factor", interestingFactors))

interestingFactors <- c(10, 8)
methodName = "JIVE"
fseaplots <- lapply(interestingFactors, function(factor) {
  plotEnrichmentUsingSeparateWeightsAndFactors(
    gseaResults[[methodName]],
    methodName,
    factor = factor,
    alpha = pval.thr,
    max.pathways = 10 # The top number of pathways to display
  )
})

cowplot::plot_grid(fseaplots[[1]], fseaplots[[2]],
                   ncol = 1, labels = paste("Factor", interestingFactors))

interestingFactors <- 4:5
methodName = "MCIA"
fseaplots <- lapply(interestingFactors, function(factor) {
  plotEnrichmentUsingSeparateWeightsAndFactors(
    gseaResults[[methodName]],
    methodName,
    factor = factor,
    alpha = pval.thr,
    max.pathways = 10 # The top number of pathways to display
  )
})

cowplot::plot_grid(fseaplots[[1]], fseaplots[[2]],
                   ncol = 1, labels = paste("Factor", interestingFactors))
```

#### Question: Explore the pathways associated with other factors and see how many shared pathways are found between different methods as well as within different factors for each method?

We observe that some of the factors are not associated with any biological pathway at all. We also observe that some factors are associated with common biological pathways, thereby, suggesting that factors are not capturing distinct biological information.

### Perform selectivity analysis

To quantify the independence of each factor in representing biological information, a selectivity score was used by Laura et al. given by:

$$Selectivity\;Score = 2 (\frac{N_c+N_f}{L})$$
Where N<sub>c</sub> is the total number of clinical annotations associated with atleast one factor, N<sub>f</sub> is the total number of factors accosiated with at least one clinical annotation, and L is the total number of associations between clinical anotations and factors.

#### Extract the selectivity results for each method
```{r}
source("../scripts/HelperForTheCourse.R")
out_bio <- biologicalComparisonModifiedWithMofaCode(weightsList, factorsList, CLL_data, pathToFseaScript, database = reactomeGS, pval.thr=pval.thr)
```

```{r}
out_bio
```

#### Question: Which method has more non-zero factors without compromising for selectivity?

#### Plot the results
```{r}
# Prepare the data frame
biological_analysis <- data.frame(
    matrix(data = NA, ncol=4, nrow=0, 
           dimnames = list(c(), c("methods", "selectivity", "nonZeroFacs", "total_pathways"))
          ),
    stringsAsFactors = FALSE)
biological_analysis <- rbind(biological_analysis,
                                data.frame(methods=rownames(out_bio), out_bio))

# Plot the selectivity score vs N_f
min_nonZero = min(biological_analysis[, "nonZeroFacs"]) 
max_nonZero = max(biological_analysis[, "nonZeroFacs"]) 
g <- ggplot(biological_analysis, 
            aes(x=nonZeroFacs,y=selectivity)) + 
    geom_point(aes(colour = methods), size=5, alpha=.6, position=position_jitter(h=0, w=0.15))+ 
    theme_bw() + 
    scale_shape_manual(values=c(15,17,16)) + 
    scale_color_manual(values=c('#FF00FF', '#FF6E28', '#C8961E', '#FF0000', '#0000FF', '#A0A0A0', '#48D1CC', '#00FF00')) +
    #ylim(floor(min((biological_analysis[,"selectivity"]*10)-.4)) / 10,
    #     ceiling(max((biological_analysis[,"selectivity"]*10)+.2)) / 10) +
    labs(title="Biological annotations", 
         x="# weights (factors) enriched in at least one annotation") +
    theme(plot.title = element_text(size=14,face="bold"),
          axis.text = element_text(size=11),
          axis.title = element_text(size=13),
          legend.text=element_text(size=10)) +
    ylab("Selectivity") + 
    ylim(c(0, 1))+
    labs(colour = "Methods", shape = "Cancer") +
    guides(color = guide_legend(order = 1),size = guide_legend(order = 3)) + 
    #scale_x_discrete() + scale_x_discrete(limits=min_nonZero:max_nonZero) + 
    scale_x_discrete(limits=min_nonZero:max_nonZero, labels = c(min_nonZero:max_nonZero));
g
```

### Print session info
```{r}
sessionInfo()
```