---
title: "Session2_2"
author: "Manik Garg"
date: "26/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Train different methods and save results

### Make sure to point towards correct python binary
```{r}
reticulate::use_python("/usr/bin/python3", required = TRUE)
```

### Load required packages
```{r}
suppressMessages(library("MOFA2"))
suppressMessages(library("MOFAdata"))
suppressMessages(library("r.jive"))
suppressMessages(library("omicade4"))
suppressMessages(library("zoo"))
```

### Load the required data
```{r}
utils::data("CLL_data")       
lapply(CLL_data,dim)
```

### Visualize the data
```{r}
#Create MOFA object
MOFAobject <- create_mofa(CLL_data)
```

```{r}
MOFAobject
```

```{r}
plot_data_overview(MOFAobject)
```

First of all we see that this dataset has both discrete (such as Mutations data) and continuous data (such as mRNA data). 
Next we see that only Mutations data has information for all the 200 patients. Therefore, this dataset does not have equal number of features for all the samples. 

Now, let us check for missing values within each data modality.

```{r}
lapply(CLL_data, function(g) table(is.na(g)))
```

We see that all of them have >= 2% of missing values. Hence, this dataset provides an opportunity to benchamrk the methods on all these real-world scenarios.

Let us start with training the MOFA model which doesn't require an explicit imputation of the missing data.

### MOFA

#### Load pre-trained MOFA model 
```{r}
MOFAobject <- readRDS(url("http://ftp.ebi.ac.uk/pub/databases/mofa/cll_vignette/MOFA2_CLL.rds"))
```

```{r}
MOFAobject
```

We limit the number of factors to 10 in all the subsequent analysis.
```{r}
numFactors = 10
```

#### Extract weights
```{r}
MOFAweights <- get_weights(MOFAobject)
```

#### Extract factors
```{r}
MOFAfactorsList <- get_factors(MOFAobject)
MOFAfactors <- MOFAfactorsList[["group1"]][, 1:numFactors]
```

### Impute the data for rest of the analysis
############# Replacing NAs with feature-wise mode in case of mutation data and feature-wise mean in case of other non-binary data ######################

```{r}
source("../scripts/HelperForTheCourse.R")

#To make the comparisons with MOFA, fill in the missing values with feature-wise mean
#Participants can play around here with various imputation techniques
CLL_data_imputed <- list("Mutations" = t(apply(CLL_data[["Mutations"]], 1, function(x) {
  x[which(is.na(x))] <- calculateMode(x, na.rm = TRUE)
  return(x)})))
CLL_data_imputed[c("Drugs", "mRNA", "Methylation")] <- lapply(CLL_data[c("Drugs", "mRNA", "Methylation")], function(x) t(na.aggregate(t(x))))
```

Note that the following sections are derived from the momix pipeline[https://github.com/ComputationalSystemsBiology/momix-notebook] made available by Cantini *et al.* (2021)[https://doi.org/10.1038/s41467-020-20430-7]

### JIVE

#### Train the model
```{r}
#Same ranks are used for the joint matrix and the individual matrices as the number of factors
#As the ranks are already specified, method="given" is used.
#conv="default" is used
#maxiter=100 is used in the interest of time.
#Feel free to explore various parameter options from the r.jive manual
JIVEfactorization <- jive(CLL_data_imputed, rankJ=numFactors, rankA = rep(numFactors, length(CLL_data_imputed)), method = "given", conv = "default", maxiter = 100, showProgress=TRUE)
```

#### Extract weights
```{r}
rankJV <- JIVEfactorization$rankJ
rankIV.v <- JIVEfactorization$rankA
J <- numeric(0)
ng <- 0
JIVEweights <- list()
for(j in 1:length(CLL_data_imputed)){
  J <- rbind(J,JIVEfactorization$joint[[j]])
  ng <- c(ng,dim(JIVEfactorization$joint[[j]])[1])
}

#Extract weights and factors from the singlular value decomposition (svd) of joint matrix J.
svd.o <- svd(J)
jV <- svd.o$v %*% diag(svd.o$d)
for(j in 1:length(CLL_data_imputed)){
  JIVEweights[[j]] <- svd.o$u[(1+sum(ng[1:j])):sum(ng[1:j+1]),1:rankJV]
  rownames(JIVEweights[[j]]) <- rownames(CLL_data_imputed[[j]])
  colnames(JIVEweights[[j]]) <- paste0("Factor", 1:numFactors)
}
names(JIVEweights) <- names(CLL_data_imputed)

```  

#### Extract factors
```{r}
JIVEfactors <- jV[,1:rankJV]
rownames(JIVEfactors) <- colnames(CLL_data_imputed[[1]])
colnames(JIVEfactors) <- paste0("Factor", 1:numFactors)
```
  
### MCIA

#### Train the model
```{r}
#Again feel free to explore more parameter options from the omicade4 reference manual
#More description in the list of outputs can be found here: https://rdrr.io/cran/ade4/man/mcoa.html
MCIAfactorization <- mcia(CLL_data_imputed, cia.nf = numFactors)
```

#### Extract weights
```{r}
MCIAweights <- list()
for(j in 1:length(CLL_data_imputed)){
    MCIAweights[[j]]<-as.matrix(MCIAfactorization$mcoa$axis[1:dim(CLL_data_imputed[[j]])[1],])
    rownames(MCIAweights[[j]])<-rownames(CLL_data_imputed[[j]])
    colnames(MCIAweights[[j]])<-paste0("Factor", 1:numFactors)
}
names(MCIAweights) <- names(CLL_data_imputed)
```

#### Extract factors
```{r}
MCIAfactors <- as.matrix(MCIAfactorization$mcoa$SynVar)
colnames(MCIAfactors) <- paste0("Factor", 1:numFactors)
```

### Compare the results

```{r}
plot_variance_explained(MOFAobject, factors = paste("Factor",1:numFactors, sep=""))
plot(JIVEfactorization)
plot(MCIAfactorization, axes = c(1,2), sample.lab = FALSE, df.color = 1:4)
```
In case of MOFA => Along with variance explained per view, variance explained per factor per view can also be visualized.
In case of JiVE => Only variance explained per view can be visualized
In case of MCIA => factors = PCs, sorted by their eigen-values as we have selected the first 10 eigenvectors for this analysis. The variance explained per factor per view is not available but overall the eigen-vectors are sorted by their eigenvalues.

### Save the imputed data and the results

```{r}
save(CLL_data_imputed, file = "../results/CllDataImputed.Rdata")
```

```{r}
weightsList = list("MOFA" = MOFAweights, "JIVE" = JIVEweights, "MCIA" = MCIAweights)
save(weightsList, file = "../results/WeightsList.Rdata")
```

```{r}
factorsList = list("MOFA" = MOFAfactors, "JIVE" = JIVEfactors, "MCIA" = MCIAfactors)
save(factorsList, file = "../results/FactorsList.Rdata")
```

### Print session info
```{r}
sessionInfo()
```