---
title: "Session2_2"
author: "Manik Garg"
date: "26/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load required packages

```{r}
suppressMessages(library("MOFA"))
suppressMessages(library("MOFAdata"))
suppressMessages(library("r.jive"))
suppressMessages(library("omicade4"))
suppressMessages(library("zoo"))
```

```{r}
source("../scripts/HelperForTheCourse.R")
```

Load the data
```{r}
data("CLL_data", package = "MOFAdata")
data("CLL_covariates", package = "MOFAdata")
```
CLL_covariates only has gender and disgnosis information

## MOFA

Create MOFA object
```{r}
MOFAobject <- createMOFAobject(CLL_data)
```

```{r}
MOFAobject
```

```{r}
plotDataOverview(MOFAobject)
```

Load pre-trained MOFA model (assuming that training has already been covered in the talk)
```{r}
filepath <- system.file("extdata", "CLL_model.hdf5", package = "MOFAdata")
MOFAobject <- loadModel(filepath, MOFAobject)
```

```{r}
MOFAobject
```
We limit the number of factors to 10 in all the subsequent analysis.
```{r}
numFactors = 10
```

Extract weights and factors matrix
```{r}
MOFAweights <- getWeights(MOFAobject)
```

```{r}
MOFAfactors <- getFactors(MOFAobject)
```

## Impute the data
############# Replacing NAs with feature-wise mode in case of mutation data and feature-wise mean in case of other non-binary data ######################
############# Or remove mutation data as imputing is too hard/dangerous ######################
```{r}
#To make the comparisons with MOFA, fill in the missing values with feature-wise mean
#Participants can play around here with various imputation techniques
CLL_data_imputed <- list("Mutations" = t(apply(CLL_data[["Mutations"]], 1, function(x) {
  x[which(is.na(x))] <- calculateMode(x, na.rm = TRUE)
  return(x)})))
CLL_data_imputed[c("Drugs", "mRNA", "Methylation")] <- lapply(CLL_data[c("Drugs", "mRNA", "Methylation")], function(x) t(na.aggregate(t(x))))
```

# Note that the following sections are derived from the momix pipeline made available by Laura et al. (2020)

## JIVE

```{r}
JIVEfactorization <- jive(CLL_data_imputed, rankJ=numFactors, rankA = rep(numFactors, length(CLL_data_imputed)), method = "given", conv = "default", maxiter = 100, showProgress=TRUE)
```

```{r}
rankJV <- JIVEfactorization$rankJ;
rankIV.v <- JIVEfactorization$rankA;
J <- numeric(0)
ng <- 0
JIVEweights <- list();
for(j in 1:length(CLL_data_imputed)){
  J <- rbind(J,JIVEfactorization$joint[[j]]);
  ng <- c(ng,dim(JIVEfactorization$joint[[j]])[1])
}
svd.o <- svd(J);
jV <- svd.o$v %*% diag(svd.o$d);
for(j in 1:length(CLL_data_imputed)){
  JIVEweights[[j]] <- svd.o$u[(1+sum(ng[1:j])):sum(ng[1:j+1]),1:rankJV]; ###error in dimension
  rownames(JIVEweights[[j]]) <- rownames(CLL_data_imputed[[j]])
  colnames(JIVEweights[[j]]) <- paste0("LF", 1:numFactors)
}
names(JIVEweights) <- names(CLL_data_imputed)

```  

```{r}
JIVEfactors <- jV[,1:rankJV]
rownames(JIVEfactors) <- colnames(CLL_data_imputed[[1]])
colnames(JIVEfactors) <- paste0("LF", 1:numFactors)
```
  
## MCIA

```{r}
MCIAfactorization <- mcia(CLL_data_imputed, cia.nf = numFactors)
```

```{r}
MCIAfactors <- as.matrix(MCIAfactorization$mcoa$SynVar)
colnames(MCIAfactors) <- paste0("LF", 1:numFactors)
```

```{r}
MCIAweights <- list()
for(j in 1:length(CLL_data_imputed)){
    MCIAweights[[j]]<-as.matrix(MCIAfactorization$mcoa$axis[1:dim(CLL_data_imputed[[j]])[1],])
    rownames(MCIAweights[[j]])<-rownames(CLL_data_imputed[[j]])
    colnames(MCIAweights[[j]])<-paste0("LF", 1:numFactors)
}
names(MCIAweights) <- names(CLL_data_imputed)
```

## Compare the results

```{r}
MOFA::plotVarianceExplained(MOFAobject)
plot(JIVEfactorization)
plot(MCIAfactorization, axes = c(1,2), sample.lab = FALSE, df.color = 1:4)
plot(MCIAfactorization, axes = c(1,3), sample.lab = FALSE, df.color = 1:4)
plot(MCIAfactorization, axes = c(2,3), sample.lab = FALSE, df.color = 1:4)
```
In case of MOFA => Along with variance explained per view, variance explained per factor per view can also be visualized.
In case of JiVE => Only variance explained per view can be visualized
In case of MCIA => factors = PCs, sorted by their eigen-values as we have selected the first 10 eigenvectors for this analysis. The variance explained per factor per view is not available but overall the eigen-vectors are sorted by their eigenvalues.

## Save the imputed data and the results

```{r}
save(CLL_data_imputed, file = "../results/CllDataImputed.Rdata")
```

```{r}
weightsList = list("MOFA" = MOFAweights,"MCIA" = MCIAweights, "JIVE" = JIVEweights)
save(weightsList, file = "../results/WeightsList.Rdata")
```

```{r}
factorsList = list("MOFA" = MOFAfactors,"MCIA" = MCIAfactors, "JIVE" = JIVEfactors)
save(factorsList, file = "../results/FactorsList.Rdata")
```

```{r}
sessionInfo()
```