---
title: "Session3_1"
author: "Manik Garg"
date: "26/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Benchmark 1: Clinical association with IGHV and trisomy 12

### Load required packages
```{r}
suppressMessages(library("MultiAssayExperiment"))
suppressMessages(library("BloodCancerMultiOmics2017"))
suppressMessages(library("pheatmap"))
suppressMessages(library("ggplot2"))
```

### Load the required data
```{r}
# Load weights obtained from each method in Session 2.1 
load("../results/WeightsList.Rdata")
# Load factors obtained from each method in Session 2.1 
load("../results/FactorsList.Rdata")
```

```{r}
# Load the CLL clinical data published by Dietrich et al. in the Bioconductor package BloodCancerMultiOmics2017
data(lpdAll, patmeta, drugs, package = "BloodCancerMultiOmics2017")
```

### Extract the relevant clinical data for this group project
```{r}
source("./HelperForTheCourse.R")
clinicalData <- extractMetadata(lpdCLL, patmeta, drugs)
```

### Only keep those samples for which relevant clinical data is available for further analysis
```{r}
factorsList = lapply(factorsList, function(g) {
  indices = match(rownames(clinicalData), rownames(g))
  return(g[indices, ])
})
```

### Save the results for later use
```{r}
save(clinicalData, file = "../results/ClinicalData.Rdata")
save(factorsList, file = "../results/UpdatedFactorsList.Rdata")
```

### Visualize the weights given to each mutation by all the factors for the specified method

```{r}
source("../scripts/HelperForTheCourse.R")
visualizeWeightsHeatmap(weightsList[["MOFA"]], 'Mutations')
visualizeWeightsHeatmap(weightsList[["JIVE"]], 'Mutations')
visualizeWeightsHeatmap(weightsList[["MCIA"]], 'Mutations')
```

### Plot top weighted mutations for a specified factor obtained from a specified method

For each factor of the specified method, the top weighted mutations can be viewed, thus providing a better way to identify factors associated with mutations of interest, especially when heatmaps become complex to interpret like in the case of MCIA and JIVE. 

For starters, lets look at MOFA factor 1 which seems to have the highest (-)ve weight for IGHV.

```{r}
methodName = "MOFA"
factorNumber = 1
nMutationsToDisplay = 10
plotTopWeightsUsingSeparateWeightsAndFactors(weightsList[[methodName]], "Mutations", factor = factorNumber, nfeatures = nMutationsToDisplay)
```
IGHV shows up at the top. 

#### Similarly explore different factors for different methods and see which factors better associate with IGHV and trisomy12 status

#### Plot sample scatter-plot

To confirm whether your identified factors can successfully divide patients into two separate groups based on their IGHV and trisomy12 status, we will plot the sample scatter. 

```{r}
IGHV <- factor(clinicalData$IGHV)
trisomy12 <- factor(clinicalData$trisomy12)
methodName <- "MOFA"
ggplot(data.frame(factorsList[[methodName]]), aes(x = #Your chosen factor related to IGHV
                                                    , y = #Your chosen factor related to trisomy12
                                                    , colour = IGHV, shape = trisomy12)) +
  geom_point()+ 
  ggtitle(methodName)
```
```{r}
methodName <- "JIVE"
ggplot(data.frame(factorsList[[methodName]]), aes(x = #Your chosen factor related to IGHV
                                                    , y = #Your chosen factor related to trisomy12
                                                    , colour = IGHV, shape = trisomy12)) +
  geom_point()+ 
  ggtitle(methodName)
```

```{r}
methodName <- "MCIA"
ggplot(data.frame(factorsList[[methodName]]), aes(x = #Your chosen factor related to IGHV
                                                    , y = #Your chosen factor related to trisomy12
                                                    , colour = IGHV, shape = trisomy12)) +
  geom_point()+ 
  ggtitle(methodName)
```
#### Question: Which method gives the best seperation between the patients having IGHV and trisomy12 mutations?

### Perform selectivity analysis

We also noticed that in JIVE, multiple factors have given highest weights to IGHV and trisomy12 status, thereby implying that the factors are capturing similar clinical information and are not exclusive to each other. To quantify the independence of each factor in representing biological information, a selectivity score was devised by Laura et al. (2020) given by:

$$Selectivity\;Score = 2 (\frac{N_c+N_f}{L})$$

Where N<sub>c</sub> is the total number of clinical annotations associated with atleast one factor, N<sub>f</sub> is the total number of factors accosiated with at least one clinical annotation, and L is the total number of associations between clinical anotations and factors.

#### Extract the selectivity results for each method

```{r}
clinicalFeaturesOfInterest <- c("IGHV", "SF3B1", "NOTCH1", "BRAF", "TP53", "del17p13", "del11q22.3", "trisomy12")
pval.thr <- 0.05
out_clinical <- calculateSelectivityScoreForClinicalAnnotations(factorsList, clinicalData, clinicalFeaturesOfInterest, pval.thr)
```

```{r}
out_clinical
```

#### Question: Which method has more non-zero factors without compromising for selectivity?

#### Plot the results
```{r}
# Prepare the data frame
clinical_analysis <- data.frame(
    matrix(data = NA, ncol=4, nrow=0, 
           dimnames = list(c(), c("methods", "selectivity", "nonZeroFacs", "total_annotations"))
          ),
    stringsAsFactors = FALSE)
clinical_analysis <- rbind(clinical_analysis,
                                data.frame(methods=rownames(out_clinical), out_clinical))

min_nonZero = min(clinical_analysis[, "nonZeroFacs"]) 
max_nonZero = max(clinical_analysis[, "nonZeroFacs"]) 
g <- ggplot(clinical_analysis, 
            aes(x=nonZeroFacs, y=selectivity)) + 
    geom_point(aes(colour = methods), size=5, alpha=.6, position=position_jitter(h=0, w=0.15))+ 
    theme_bw() + 
    #scale_shape_manual(values=c(15,17,16)) + 
    scale_color_manual(values=c('#FF00FF', '#FF6E28', '#C8961E', '#FF0000', '#0000FF', '#A0A0A0', '#48D1CC', '#00FF00')) +
    ylim(floor(min((clinical_analysis[,"selectivity"]*10)-.4)) / 10,
         ceiling(max((clinical_analysis[,"selectivity"]*10)+.2)) / 10) +
    labs(title="Clinical annotations", 
         x="# metagenes (factors) enriched in at least one annotation") +
    theme(plot.title = element_text(size=14,face="bold"),
          axis.text = element_text(size=11),
          axis.title = element_text(size=13),
          legend.text=element_text(size=10)) +
    ylab("Selectivity") + 
    labs(colour = "Methods",shape="Cancer") +
    guides(color = guide_legend(order = 1), size = guide_legend(order = 3)) + 
    #scale_x_discrete() + scale_x_discrete(limits=min_nonZero:max_nonZero) + 
    scale_x_discrete(limits=min_nonZero:max_nonZero, labels = c(min_nonZero:max_nonZero));
g
```

### Print session info
```{r}
sessionInfo()
```